Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2024-08-21 analysis/paper/paper.qmd
work on the methods

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest settlement in northeastern Taiwan [@Chen2007]. Kiwulan encountered European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. What was the resource use of Indigenous people before and after the foreign presence? Can we identify shifts in these practices in response to colonial economic demands? To address these questions, I will focus on subsistence practices because the choices made in the practices can reveal the strategies Indigenous people used to respond to European exploitation with associated changes in their environment. I will primarily examine the zooarchaeological evidence from Kiwulan for subsistence practices, coupled with ceramic residues for culinary practices. Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

# European encounters in the northern Taiwan
<!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp. 103-134; @Berrocal2018]. The neighboring area of the fort was inhavited by local indigenous groups, the Taparris and the Quimaurris, who had lived at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The relations between the Spanish and indigenous people might not involve extensive colonialism or intercultural domination, and the Spanish territorial control could be limited to the fort. Instead, the Spanish relied on the exchange of local goods and food surpluses with local indigenous people, according to written records [@Borao2001, SIT:170-171]. This corresponds with a higher proportion of indigenous and Chinese items with little European material found in San Salvador [@Berrocal2018], indicating cultural entanglement.

@@

# Archaeological context at Kiwulan
<!--introduction of Kiwulan and the population-->
Kiwulan, located in the northern part of Yilan, is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. As a salvage project, a large scale excavation was conducted at Kiwulan to recover information before it was destroyed. A total of 262 excavation units uncovered abundant artifacts and features, including middens, burials, and post-holes with wooden posts in situ. It suggests that Kiwulan is a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. The main inhabitants of Kiwulan were the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

@@

# Mateirals and methods
To explore changes in the subsistence strategies before and after the European arrival with associated long-distance trade network, I examined zooarchaeological and botanical evidence related to economic activities in a colonial context from Kiwulan. The faunal and botanical assemblages analyzed in this study are from 40 units located in the residential area excavated during the first season (2001-2006) at Kiwulan. The 40 excavation units were selected due to their relativity intact chronological contexts and spatially continuous area. Because deerskin was the main commodity in local regional exchange since the 14th century and further largely demanded during the Dutch colonization period, deer bones and their butchery patterns are especially focused.

>>>>

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest settlement in northeastern Taiwan [@Chen2007]. Kiwulan encountered European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. What was the resource use of Indigenous people before and after the foreign presence? Can we identify shifts in these practices in response to colonial economic demands? To address these questions, I will focus on subsistence practices because the choices made in the practices can reveal the strategies Indigenous people used to respond to European exploitation with associated changes in their environment. I will primarily examine the zooarchaeological assemblages from Kiwulan for subsistence practices, coupled with botanical evidence as a supplementary evidence. Zooarchaeology can offer insights into how animals were utilized in response to economic strategies within certain social contexts, such as intercultural contact or European colonization (DeFrance, 2009). Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

# European Encounters in the Northern Taiwan
<!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp. 103-134; @Berrocal2018]. The neighboring area of the fort was inhavited by local indigenous groups, the Taparris and the Quimaurris, who had lived at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The relations between the Spanish and indigenous people might not involve extensive colonialism or intercultural domination, and the Spanish territorial control could be limited to the fort. Instead, the Spanish relied on the exchange of local goods and food surpluses with local indigenous people, according to written records [@Borao2001, SIT:170-171]. This corresponds with a higher proportion of indigenous and Chinese items with little European material found in San Salvador [@Berrocal2018], indicating cultural entanglement.

@@

# Archaeological Context at Kiwulan
<!--introduction of Kiwulan and the population-->
Kiwulan, located in the northern part of Yilan, is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. As a salvage project, a large scale excavation was conducted at Kiwulan to recover information before it was destroyed. A total of 262 excavation units uncovered abundant artifacts and features, including middens, burials, and post-holes with wooden posts in situ. It suggests that Kiwulan is a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. The main inhabitants of Kiwulan were the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

@@

# Mateirals and Methods
To explore changes in resource use before and after the European arrival with associated long-distance trade, I analyzed zooarchaeological and botanical assemblages from 40 adjacent units at the residential area excavated during the first season (2001-2004) at Kiwulan [@Chen2007]. These 40 units were selected based on their relatively intact archaeological contexts and spatial continuity, providing a consistent framework for understanding daily food practices throughout different time periods on a settlement scale.

To construct a reliable chronology, the archaeological records were divided into six temporal layers spanning from the 14th century to the 19th century based on excavation depth, stratigraphic data, sediment color, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above still provides a feasible framework for analyzing archaeological records before and after the European influence. Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with Chinese migrants.

In this study, only the faunal assemblages identified to at least the family level were considered for consistent comparison across time sequences. To obtain a general pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. In this analysis, ubiquity values are expressed as the percentage of units with a given taxon to the total number of units containing cultural layers. Taxonomic abundance was quantified using the Number of Identified Specimens (NISP) method for a broad overview of taxa through time [@Grayson2004; @Lyman2012]. For botanical assemblages, their number was served as supplementary evidence for food practices due to potential biases that caused by differential preservation conditions. For example, carbonized seeds usually preserve better than uncarbonized ones, which could lead to a misleading interpretation.

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the skeletal compositions and butchery patterns of deer bones were examined. The distribution of body portions and butcher patterns can shed light on how indigenous people utilize deer in response to economic demands during the foreign presence. The proportion of deer skeletal elements was quantified using the Number of Identified Specimens (NISP) and classified into three groups: meaty elements (e.g., upper limb), mid-meaty elements (e.g., body parts), and non-meaty elements (e.g., cranial parts and antlers) [@Madrigal2004]. For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified. The proportion of deer bones with cut marks was calculated using the normed NISP for comparison between temporal sequences [@Grayson2004]. In addition, cut marks associated with the skinning process were examined. The skinning process can be identified by the encircling marks left around the lower limb, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].
